Value Ordering, Discrepancies, and Restarts for Subgraph Algorithms\thanks{This footnote leaves room for a piece of text saying how grateful we are to our funding agency. This includes a magic grant number, so it takes up three precious lines of space.}

### 2018-01-29 paper/paper.tex
Are we nearly there yet?

\documentclass{article}
\usepackage{ijcai17}
\usepackage{times}
\usepackage{microtype}
\usepackage{amsmath}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage{tikz}

@@

\title{Value Ordering, Discrepancies, and Restarts for Subgraph Algorithms\thanks{This work was
supported by the Engineering and Physical Sciences Research Council [grant numbers EP/026842/1 and EP/M508056/1]}}
\author{Ciaran McCreesh \and Patrick Prosser \and James Trimble\\ University of Glasgow, Glasgow, Scotland \\
    ciaran.mccreesh@glasgow.ac.uk}

@@

Let $G$ be a graph. We denote its set of vertices by $\vertexset(G)$, and write $v \sim_G w$ to mean
that vertices $v$ and $w$ are adjacent; we allow a vertex to be adjacent to itself. The
\emph{neighbourhood} of a vertex, $\neighbourhood_G(v)$, is the set of vertices adjacent to $v$, and
the \emph{degree} of $v$, $\deg_G(v)$, is the cardinality of its neighbourhood; when the graph is
clear from the context, we omit the $G$ subscript. The \emph{neighbourhood degree sequence} of $v$,
$\nds_G(v)$, is the sequence of degrees of the neighbours of $v$ in descending order, and we denote
by $\preceq$ the inclusion relation of \citet{DBLP:journals/constraints/ZampelliDS10}. As per
\citet{DBLP:conf/aaai/HoffmannMR17}, $G^{n,\ell}$ denotes the graph with vertex set $\vertexset(G)$,
and with edges between vertices $v$ and $w$ if there are at least $n$ simple paths of length exactly
$\ell$ between vertices $v$ and $w$ in $G$.

@@

Another alternative to plain backtracking search is provided by \emph{restarts}. The general idea is
to perform a certain amount of search, and then if no solution has been found (and unsatisfiability
has not been proven), to abandon search and restart from the beginning. Obviously, such an approach
can only be beneficial if something changes after restarting---in a constraint programming setting,
this is usually the variable-ordering heuristic
\citep{DBLP:journals/jsat/LecoutreSTV07,DBLP:conf/cp/GayHLS15,DBLP:conf/aaai/LeeSZ16,DBLP:conf/cp/GlorianBLLM17}.
In this paper, we instead rely upon randomness in the \emph{value}-ordering heuristic, and keep the
variable ordering fixed.\footnote{It may be possible to further improve \cref{algorithm:sip} by also
introducing randomness or some form of learning into its variable-ordering heuristic. However,
introducing a second change would considerably complicate the empirical analysis. We therefore leave
this change as future work.} Using restarts on value-ordering heuristics appears to be uncommon,
although \citet{DBLP:conf/flairs/RazgonOP07} look at learning value-ordering heuristics from
restarts, and \citet{DBLP:conf/cp/ChuSS09} use a similar scheme in the context of parallel search.

@@

We use the Luby scheme \citep{DBLP:journals/ipl/LubySZ93} to determine when to restart. We count the
number of backtracks (that is, when we reach the end of the main $\KwSty{foreach}$ loop on
\siplineref{countrestarts}) to decide when to restart, rather than counting recursive calls.
Following established wisdom, we multiply each item in the Luby sequence by a magic constant.
Preliminary experiments demonstrated that a constant 100 is slightly too small and 1,000 is slightly
too large, but lacking a more principled way of selecting the constant's value, we fell back on
divine revelation and use a multiplier of 666. (We also tried a geometric scheme. The results were
less favourable---it appears that value-ordering heuristics benefit from a very aggressive restart
strategy.)

The remainder of the changes are to introduce nogoods. To avoid exploring portions of the search
space that we have already visitied, every time we restart, we add new constraints to the problem
which eliminate already-explored subtrees. We generate simple decision nogoods, as follows. The $B$
argument to $\FuncSty{search}$ tracks the branching choices made so far, in the form of a CNF clause
with each literal being a guessed $(v \mapsto v')$ assignment. Upon backtracking due to a decision to
restart, we post a nogood of the form $(v \mapsto v') \wedge (w \mapsto w') \wedge (x \mapsto x')
\Rightarrow \bot$ for the current branch, and for every branch to its left in the search tree.

@@

In \cref{figure:restarts} we show the effects of adding restarts to the algorithm. With restarts,
the random value-ordering heuristic comfortably beats the degree strategy without restarts. In other
words, although having a good variable-ordering heuristic is beneficial, introducing randomness into
the search is better, if it is done alongside a mechanism to avoid heavy commitment to any
particular random choice. \Cref{figure:scatter-random} looks at this in more detail. The performance
on unsatisfiable instances, drawn as dark dots, remains unchanged as expected (which shows that the
overheads of nogoods and restarting are negligible). However, although better in aggregate on
satisfiable instances, on a case by case basis many instances perform substantially worse with
randomisation and restarts. ?? Quantify this a bit

@@

\Cref{figure:restarts} shows that the biased heuristic together with restarts is better still---that
is, if we are introducing restarts, then it is better to add a small amount of randomness to a
tailored deterministic heuristic than it is to throw away the heuristic altogether.

In the more detailed view in \cref{figure:scatter-by-family}, comparing the basic algorithm with the
degree heuristic to the degree-biased algorithm with restarts, all of the unsatisfiable instances
are very close to the $x-y$ diagonal, showing that their performance is nearly unchanged. On the
other hand, there are ?? large numbers of satisfiable instances well below the diagonal line,
indicating large speedups.  Better yet, there are only ?? satisfiable instances that are more than a
factor of ten times worse. In other words, as well as improving performance, we have made up most of
the consistency we lost by introducing randomness.

@@

Two recent maximum common induced subgraph algorithms also make use of backtracking search with
degree as a value-ordering heuristic. The k${\downarrow}$ algorithm
\citep{DBLP:conf/aaai/HoffmannMR17} attempts to solve the problem by first trying to solve the
induced subgraph isomorphism problem, and then if that fails, retries allowing a single unmatched
vertex (and thus using weaker invariants), and so on. Due to its similarity to \cref{algorithm:sip},
we can introduce the same bias and restart strategy, requiring only a small amount of jiggery-pokery
to handle wildcards in nogoods.  Meanwhile, the McSplit algorithm
\citep{DBLP:conf/ijcai/McCreeshPT17} uses a constraint programming style search, but with special
propagators and backtrackable data structures that exploit special properties of the problem. The
unconventional domain store used by McSplit precludes the use of arbitrary unit propagation, and so
when introducing restarts, we cannot propagate using nogoods.  Instead, we can only detect when we
are inside an already-visited branch.  We must therefore use the one watched literal scheme instead.

?? Benchmark results on the same families of instances are shown in
\cref{figure:kdown,figure:mcsplit,figure:mcs}. Clearly beneficial. Note that on McSplit we're doing
extra work and this shows, but it's still worthwhile.

\begin{figure}[p]
    \centering
    \includegraphics*{gen-graph-scatter-kdown.pdf}
    \caption{k${\downarrow}$ speedups}\label{figure:kdown}
\end{figure}

\begin{figure}[p]
    \centering
    \includegraphics*{gen-graph-scatter-mcsplit.pdf}
    \caption{McSplit speedups}\label{figure:mcsplit}
\end{figure}

\begin{figure}[p]
    \centering
    \includegraphics*{gen-graph-mcs.pdf}
    \caption{k${\downarrow}$ and McSplit cumulative curves}\label{figure:mcs}
\end{figure}

\section{Conclusion}

?? This stuff is quite good.

?? We can keep good value-ordering heuristics, and get large improvement. Importantly, we can do
this with almost no effect on unsatisfiable instances, and rare to make satisfiable much worse.

?? No theoretical justification for biased. Also tried polynomial weightings, exponential using
$\left|D_v\right|$ as the base. Using $e$ (or for that matter, $2$ or $\pi$) as a base seems to be
``just the right amount'' of randomness.

?? Variable ordering

?? Bigger CP picture.

?? Parallel future work

>>>>

\documentclass{article}
\pdfpagewidth=8.5in
\pdfpageheight=11in
\usepackage{ijcai18}
\usepackage{times}
\usepackage{xcolor}
\usepackage{soul}
\usepackage[utf8]{inputenc}
\usepackage[small]{caption}
\usepackage{microtype}
\usepackage{amsmath}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage{tikz}

@@

\title{Value Ordering, Discrepancies, and Restarts for Subgraph Algorithms\thanks{This footnote
leaves room for a piece of text saying how grateful we are to our funding agency. This includes a
magic grant number, so it takes up three precious lines of space.}}
\author{Anonymous \and Anonymous \and Anonymous \\ Somewhere \\ Some Email Address}

@@

Let $G$ be a graph. We denote its set of vertices by $\vertexset(G)$, and write $v \sim_G w$ to mean
that vertices $v$ and $w$ are adjacent; we allow a vertex to be adjacent to itself. The
\emph{neighbourhood} of a vertex, $\neighbourhood_G(v)$, is the set of vertices adjacent to $v$, and
the \emph{degree} of $v$, $\deg_G(v)$, is the cardinality of its neighbourhood. The
\emph{neighbourhood degree sequence} of $v$, $\nds_G(v)$, is the sequence of degrees of the
neighbours of $v$ in descending order, and we denote by $\preceq$ the inclusion relation of
\citet{DBLP:journals/constraints/ZampelliDS10}. When the graph is clear from the context, we omit
the $G$ subscripts. As per \citet{DBLP:conf/aaai/HoffmannMR17}, $G^{n,\ell}$ denotes the graph with
vertex set $\vertexset(G)$, and with edges between vertices $v$ and $w$ if there are at least $n$
simple paths of length exactly $\ell$ between vertices $v$ and $w$ in $G$.

@@

Another alternative to plain backtracking search is provided by \emph{restarts}. The general idea is
to perform a certain amount of search, and then if no solution has been found (and unsatisfiability
has not been proven), to abandon search and restart from the beginning. Obviously, such an approach
can only be beneficial if something changes after restarting---in a constraint programming setting,
this is usually the variable-ordering heuristic
\citep{DBLP:journals/jsat/LecoutreSTV07,DBLP:conf/cp/GayHLS15,DBLP:conf/aaai/LeeSZ16,DBLP:conf/cp/GlorianBLLM17}.
In this paper, we instead rely upon randomness in the \emph{value}-ordering heuristic, and keep the
variable ordering fixed.\footnote{It may be possible to further improve \cref{algorithm:sip} by also
introducing randomness or some form of learning into its variable-ordering heuristic. However,
simultaneously introducing a second change would considerably complicate the empirical analysis.}
Using restarts on value-ordering heuristics appears to be uncommon, although
\citet{DBLP:conf/flairs/RazgonOP07} look at learning value-ordering heuristics from restarts, and
\citet{DBLP:conf/cp/ChuSS09} use a similar scheme in the context of parallel search.

@@

We use the Luby scheme \citep{DBLP:journals/ipl/LubySZ93} to determine when to restart. We count the
number of backtracks (that is, when we reach the end of the main $\KwSty{foreach}$ loop on
\siplineref{countrestarts}) to decide when to restart, rather than counting recursive calls.
Following established wisdom, we multiply each item in the Luby sequence by a magic constant---we
used SMAC \citep{DBLP:conf/lion/HutterHL11} to select a value of 660. (We also tried a geometric
restart scheme. The results were less favourable---it appears that value-ordering heuristics benefit
from a very aggressive restart strategy.)

The remainder of the changes are to introduce nogoods. To avoid exploring portions of the search
space that we have already visitied, every time we restart, we add new constraints to the problem
which eliminate already-explored subtrees. We generate simple decision nogoods, as follows. The $B$
argument to $\FuncSty{search}$ tracks the branching choices made so far, in the form of a CNF clause
with each literal being a guessed $(v \mapsto v')$ assignment. Upon backtracking due to a decision to
restart, we post a nogood of the form $(v \mapsto v') \wedge (w \mapsto w') \wedge (x \mapsto x')
\Rightarrow \bot$ for every branch to the left of the current (incomplete) branch in the search tree.

@@

In \cref{figure:restarts} we show the effects of adding restarts to the algorithm. With restarts,
the random value-ordering heuristic comfortably beats the degree strategy without restarts. In other
words, although having a good variable-ordering heuristic is beneficial, introducing randomness into
the search is better, if it is done alongside a mechanism to avoid heavy commitment to any
particular random choice. \Cref{figure:scatter-random} looks at this in more detail. The performance
on unsatisfiable instances, drawn as dark dots, remains unchanged as expected (which shows that the
overheads of nogoods and restarting are negligible). However, although better in aggregate on
satisfiable instances, on a case by case basis many instances perform substantially worse with
randomisation and restarts.

@@

\Cref{figure:restarts} also shows that the biased heuristic together with restarts is better
still---that is, if we are introducing restarts, then it is better to add a small amount of
randomness to a tailored deterministic heuristic than it is to throw away the heuristic altogether.
Indeed, the original algorithm can solve 1,983 satisfiable instances by 993.0 seconds, whilst the
biased and random restarting algorithms require only 9.2 seconds and 33.8 seconds respectively to
solve the same number.

In the more detailed view in \cref{figure:scatter-by-family}, comparing the basic algorithm with the
degree heuristic to the degree-biased algorithm with restarts, all of the unsatisfiable instances
are very close to the $x-y$ diagonal, showing that their performance is nearly unchanged. On the
other hand, there are large numbers of satisfiable instances well below the diagonal line,
indicating large speedups.  Better yet, there are only a handful of satisfiable instances that are
more than a factor of ten times worse. In other words, as well as improving performance, we have
made up most of the consistency we lost by introducing randomness.

@@

Two recent maximum common induced subgraph algorithms also make use of backtracking search with
degree as a value-ordering heuristic. The k${\downarrow}$ algorithm
\citep{DBLP:conf/aaai/HoffmannMR17} attempts to solve the problem by first trying to solve the
induced subgraph isomorphism problem, and then if that fails, retries allowing a single unmatched
vertex (and thus using weaker invariants), and so on. Due to its similarity to \cref{algorithm:sip},
we can introduce the same bias and restart strategy, requiring only a small amount of jiggery-pokery
to handle wildcards in nogoods.  Meanwhile, the McSplit${\downarrow}$ algorithm
\citep{DBLP:conf/ijcai/McCreeshPT17} uses a constraint programming style search, but with special
propagators and backtrackable data structures that exploit special properties of the problem. The
unconventional domain store used by McSplit${\downarrow}$ precludes the use of arbitrary unit propagation, and so
when introducing restarts, we cannot propagate using nogoods.  Instead, we can only detect when we
are inside an already-visited branch.  We must therefore use the one watched literal scheme instead.

Performance results from these two modified algorithms, using the same families of instances as in
the previous section, are shown in \cref{figure:kdown,figure:mcsplit,figure:mcs}. Although we have
moved from a decision problem to an optimisation problem, the same changes remain clearly
beneficial. For the $k{\downarrow}$ algorithm, the change has a minimal effect on many instances
(typically, where the $k = 0$ subproblem is unsatisfiable and hard, and the $k = 1$ subproblem is
satisfiable and easy), but gives large benefits on many more instances than it penalises: it is over
an order of magnitude better on nearly a hundred instances, whilst being an order of magnitude worse
on only five.

With McSplit${\downarrow}$, the inability to use two watched literals means that in many cases we
introduce a small slowdown. However, the overall pattern is the same: when introducing restarts and
a biased value ordering heuristic, it is much more common to see a large speedup than it is a large
slowdown.

\begin{figure}[p]
    \centering
    \includegraphics*{gen-graph-mcs.pdf}
    \caption{The cumulative number of maximum common induced subgraph instances solved by
    k${\downarrow}$ and McSplit${\downarrow}$ over time, with and without restarts.}\label{figure:mcs}
\end{figure}

\begin{figure}[p]
    \centering
    \includegraphics*{gen-graph-scatter-kdown.pdf}
    \caption{Comparing k${\downarrow}$ with and without restarts on an instance by instance basis.}\label{figure:kdown}
\end{figure}

\begin{figure}[p]
    \centering
    \includegraphics*{gen-graph-scatter-mcsplit.pdf}
    \caption{Comparing McSplit${\downarrow}$ with and without restarts on an instance by instance
    basis.}\label{figure:mcsplit}
\end{figure}

\section{Conclusion and Future Work}

We have shown that it is both possible and beneficial to introduce a small amount of randomness into
backtracking search algorithms for hard subgraph problems, without compromising existing search
order heuristics, and without paying a penalty for unsatisfiable instances.

The success of this approach raises a number of interesting questions and avenues for possible
future research. For example, will this same technique work for other problems where we have a good
value-ordering heuristic? Are there further benefits to be had from also modifying variable-ordering
heuristics? Are there theoretical rather than just empirical reasons for using the
\emph{softmax}-like bias scheme? And how can such a search process be parallelised?

\section*{Acknowledgments}

We would like to thank some anonymous people for their comments. These people have names that take
up approximately this amount of space, which is helpful for us getting the page layout right when
anonymising the paper for review.

% Ruth Hoffmann, Christophe Lecoutre, Christine Solnon, Craig Reilly.
